Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-11-10 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Johanna commit 2e8b414c631dc873416177dee20c7d44108de096

```{r clus,  echo = FALSE, warning = FALSE,cache = TRUE}
concepts <- data.frame(Concept=c("Sample unit","Hierarchical levels","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
                       Details=c("Units of the population from which measurements are taken in a sample.",
                                 "Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.",
                                 "Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.",
                                 "Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.",
                                 "Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).",
                                 "The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
                                 ""))
library(kableExtra)
concepts%>%
  kable( booktabs = T, escape = F,
         caption = "Concepts in multilevel methods",
         col.names = c("Concept","Details"), align = "l")%>%
  column_spec(c(1), width = "3cm")%>%
  column_spec(c(2), width = "12cm")
```

@@

Install non-CRAN packages if necessary:
```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")

```

@@

The estimated effects in the complete data are visualized in Figure \ref{}.

```{r forest, echo=FALSE}
fits <-
  purrr::map_dfr(
    unique(impact$name),
    ~ {
      glm(
        mort ~ 1 + age + motor_score + pupil + ct,
        family = "binomial",
        data = filter(impact, name == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(name = as.character(.x), .)
  ) %>% filter(term %in% c("ctIII", "ctIV/V"))

n <- impact %>% group_by(name) %>% 
  tally()

fits %>% left_join(n, by = "name") %>% 
  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
  theme_classic() +
  scale_size(guide = 'none')
# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# TODO: missingness prop inside clusters
# TODO: make row height proportional to pattern frequency
```
<!-- We will use the following estimates as comparative truth in this tutorial: -->
```{r truth}
# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

@@

To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
```

@@

```{r impact-fit, cache=TRUE}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

@@

In this example, we demonstrate a multilevel imputation of an intercept and slope random model with a continuous response. We use the obesity dataset included in the `micemd`@ package, a synthetic dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. To easy the explanations we simulate data for 5 clusters and we reduce the dataset to the following variables:

- `region` Cluster variable,
- `gender` Gender (0=male, 1=female),
- `age` Age of the patient,
- `height` Height in meters,
- `weight` Weight in kilograms,
- `time`` Response time in minutes (inclusion-restriction variable).
- `FamOb`` Family obesity history (yes or not)

In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism. This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).

In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors. Researchers, based on observable data, are considering the inclusion of random intercept effects, as well as introducing a random slope for the Age variable.

The main model is given by:
$$BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}$$
We start by loading the data:
```{r, include=FALSE}
library(here)
library(performance)
source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
load(here("heckman","Obesity.rda"))
```

```{r obesity}
#data("data_heckman", package = "micemd")
#dat <- data_heckman
```

Now, let's begin by examining the missing patterns in the data by cluster:
```{r obesity-md}
require(patchwork)
  myplots <- lapply(1:5, function(i) {
         ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
           ggtitle(paste0("Cluster", i))+ 
      theme(legend.position = ifelse(i==1,"bottom","none"))
      })
  myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]

```

We observe that the missing pattern is quite similar across the clusters. However, regarding the weight variable, we notice that is systematically missing in cluster 3.
In order to evaluate if it is required a imputation method for 1-level or 2-level we assess the Intraclass Correlation Coefficient (ICC) for the outcome variable, we use the "performance" package:

```{r obesity icc}
Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
performance::icc(Nulmodel)
```

Since the ICC is above 0.1 and as the main analyis will be consider a mixed model, we decide to use two-level (2l) imputation methods. In this imputation process, we include all predictor variables in the main model and since BMI is a composite variable. We also incorporate weight and height variables.

To determine the best imputation method for these variables, we can use the `find.defaultMethod`function provided in the `micemd` package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters. It suggests using "2l.2stage.bin" for the FAV variable and "2l.2stage.norm" for the age variable. However, after inspecting the age density plot, we consider modifying its method to "2l.2stage.pmm." For the BMI variable, we employ deterministic imputation.
```{r obsmar_met}
meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
meth_mar["Age"]<-"2l.2stage.pmm" 
```
For these imputation models, it is necessary to specify the prediction matrix, with the cluster variable labeled as -2 and the predictor variable labeled as 2, encompassing all variables. We need to suprime the variable Time as this variable is not especified in the main model and modify also in the prediction matrix the relationship between BMI, weight and height to avoid circular predictions. Then we proced to run the imputation model.

```{r obsmar_pred,echo = T}
pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
pred_mar[,"Cluster"]<- -2 # clustering variable
pred_mar[,"Time"]<-0
pred_mar[pred_mar==1]<-2
pred_mar[c("Height", "Weight"), "BMI"] <- 0
ggmice::plot_pred(pred_mar)
```
```{r obsmar_mice,echo = T,results="hide",include=FALSE}
imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar, m=10, seed = 123)
```

```{r obsmar_plot}
summary(complete(imp_mar,"long")$Weight)
plot(imp_mar)
```
We are also contemplating the utilization of the "pmm" option, given that the imputed values' weights are lower than those of the observable values.

```{r obsmarp_mice,results="hide",include=FALSE}
meth_mar["Weight"]<-"2l.2stage.pmm" 
imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar, m=10, seed = 123)
```
```{r obsmarp_plot}
summary(complete(imp_mar_pmm,"long")$Weight)
plot(imp_mar_pmm)
```
After confirming convergence, we proceed to save the results for future use. Users are considering the possibility that the weight variable may not have been selected randomly. It's likely that an omitted variable, like self-esteem, could influence this selection. For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.

To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models. Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested. In contrast, the selection model approach includes methods such as the Heckman model, which can be particularly useful in this case. Several methods, including Galimar2017, Hammon2021, and the recent Munoz method, designed for two-level data, allowing for variations in intercepts and exposure effects (random intercept and slope).

To apply the heckman method, the weight variable should be specified as "2l.2stage.heckman" found in the micemd package. Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.

Here for the outcome equation we consider is the same imputation model that we used for the MAR case (main model). 
$$Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}$$
Regarding the selection equation, we include the same predictors as those in the primary model, as well as a time variable. Here the time variable serves as a restriction exclusion variable specifically explaining the probability of being observed but not affecting the incomplete value (Weight). In this context, we assume that the time a user spends completing the survey serves as a proxy for the barriers they may encounter in survey completion, such as familiarity with the survey content or internet speed. These factors may lead the user to skip specific questions or even the entire survey. Also we assume the time is not have any influence on the weight assigned to the subject. 
$$R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$
These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution. For a more comprehensive understanding of the model and the exclusion restriction, you can refer to (Munoz).

To integrate information from both equations, we must adjust the prediction matrix. The cluster variable remains specified as before (-2). In this imputation method, all the variables present in both the selection and outcome equations are included with a random effect. However, it's essential to distinguish which of these variables appear in each equation.

In this framework, when a variable is shared between both equations, it is denoted as (2). Predictors exclusive to the outcome equation are indicated as (-4), while those exclusive to the selection equation are labeled as (-3). Consequently, the only alteration needed in the predictor matrix pertains to the variable "Time."

```{r obsmnar_pred}
pred_mnar <- pred_mar
pred_mnar["Weight","Time"]<- -3
ggmice::plot_pred(pred_mnar)
```

We also need to modify the method of the weight variable
```{r obsmnar_meth}
meth_mnar <- meth_mar
meth_mnar["Weight"]<- "3l.2stage.heckman"
```

Then we proceed to run the imputation model as before 
Impute the missingness:
```{r obsmnar_mice,results="hide",include=FALSE}
imp_mnar <- mice(
  Obesity, # dataset with missing values
  m = 10, # number of imputations
  maxit = 10,
  meth = meth_mnar, #imputation method vector
  pred = pred_mnar, #imputation predictors matrix
  meta_method = "reml",
  pred_std=FALSE,
  printFlag = FALSE
)
```

After executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values. Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive). Consequently, we are contemplating a different approach, specifically, the use of "pmm," which involves imputing values from donor observations. This approach ensures that the imputed values remain within the range of observable values.

```{r obsmnar_plot}
summary(complete(imp_mnar,"long")$Weight)
plot(imp_mnar)
```

We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by seeting the pmm parameter as true.
```{r obsmnarp_mice,results="hide",include=FALSE}
imp_mnar_pmm <- mice(
  Obesity, # dataset with missing values
  m = 10, # number of imputations
  maxit = 10,
  meth = meth_mnar, #imputation method vector
  pred = pred_mnar, #imputation predictors matrix
  meta_method = "reml",
  pred_std=FALSE,
  pmm=TRUE,
  printFlag = FALSE
)
```

```{r obsmnarp_plot}
summary(complete(imp_mnar_pmm,"long"))
plot(imp_mnar_pmm)
```

Then after this modification we proceed to compare the effects on the model, therefore we run the analysis model on each of the completed datasets along with the dataset where the incomplete values are removed (CC).

```{r models}
library(ggplot2)
cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
```

We note that there is minimal disparity in the age effect, FamObs, across the various imputation models under consideration. Nonetheless, with respect to Gender, we observe a notable distinction: in the M(C)AR assumption, the effect is significant, whereas under the MNAR assumption, it becomes insignificant. An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.

@@

Visualize missing data pattern:

@@

>>>>

Table 1: Concepts in multilevel methods

| **Concept**         | **Details**                                                                   |
|---------------------|-------------------------------------------------------------------------------|
| Sample unit         | Units of the population from which measurements are taken in a sample.        |
| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
|                     | the values observed in the study are representative of all values in the      |
|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
|                     | admission we might select only certain hospitals that are representative of   |
|                     | the difference of y between individual hospitals, but rather the variation of |
| ICC                 | The variability due to clustering is often measured by means of the           |
|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
|                     | of variance that can be attributed to the cluster-level, where a high         |
|                     | ICC would indicate that a lot of variability is due to the cluster            |
|                     | structure.                                                                    |
| Random effect       | Multilevel models typically accommodate for variability by including          |
|                     | a separate group mean for each cluster. In addition to random                 |
|                     | intercepts, multilevel models can also include random coefficients            |
|                     | and heterogeneous residual error variances across clusters [see e.g.          |
|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |

@@

@@

<!-- visualized in Figure \ref{}. -->

```{r forest, echo=FALSE}
# fits <-
#   purrr::map_dfr(
#     unique(impact$name),
#     ~ {
#       glm(
#         mort ~ 1 + age + motor_score + pupil + ct,
#         family = "binomial",
#         data = filter(impact, name == .x)
#       )
#     } %>%
#       broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
#       .[, c("term", "estimate", "conf.low", "conf.high")] %>%
#       cbind(name = as.character(.x), .)
#   ) %>% filter(term %in% c("ctIII", "ctIV/V"))
# 
# n <- impact %>% group_by(name) %>% 
#   tally()
# 
# fits %>% left_join(n, by = "name") %>% 
#   ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
#   geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
#   geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
#   geom_point(shape = 20, position = position_dodge(width = 0.5)) +
#   labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
#   theme_classic() +
#   scale_size(guide = 'none')
# # TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# # TODO: missingness prop inside clusters
# # TODO: make row height proportional to pattern frequency
```

We will use the complete data estimates as comparative truth in this tutorial. The estimated effects in the complete data are presented in Table XYZ. 
```{r impact_truth, include=FALSE, echo=TRUE}
fit <- glmer(mod, family = "binomial", data = impact) # fit the model
tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

@@

To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
```

@@

Evaluate the convergence of the algorithm:
```{r impact-conv, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
plot_trace(imp)
```

Analyze the imputed data:
```{r impact-fit, cache=TRUE}
fit <- imp %>% 
  with(glmer(
    mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
    family = "binomial"
    )) 
# tidy(pool(fit))
# as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```
The estimated effects after imputation are presented in Table XYZ.

@@

TODO: explain exclusion restriction.

Data are simulated and included in the `micemd` package. We will use the following variables:

- `region` Cluster variable,
- `gender` Gender (0=male, 1=female),
- `age` Age of the patient,
- `height` Height in meters,
- `weight` Weight in kilograms,
- `time`` Response time in minutes (inclusion-restriction variable).

The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)

@@

Load the data:
```{r obesity}
data("data_heckman", package = "micemd")
```

Load the data:
```{r}
#data("obesity", package = "miceheckman")
```

Select relevant variables:
```{r bmi}
#dat <- select(obesity, -"bmi")
```

Create predictor matrix:
```{r}
#pred <- quickpred(dat)   # predictor matrix
#pred["weight","cluster"]<- -2 # clustering variable
#pred["weight","rt"] <- -3 #inclusion-restriction variable
#plot_pred(pred)
```

Set the Heckman model as imputation method:
```{r}
#meth <- make.method(dat) # methods vector
#meth["weight"]<-"2l.heckman"
```

Visualize missing data pattern:
```{r obesity-md}
plot_pattern(dat)
```

@@

Create predictor matrix:
```{r obesity-pred, eval = FALSE}
pred <- quickpred(dat)   # predictor matrix
pred[,"Cluster"] <- -2 # clustering variable
pred[, "Weight"] <- -3 #inclusion-restriction variable
plot_pred(pred)
```

Set the Heckman model as imputation method:
```{r, eval = FALSE}
meth <- make.method(dat) # methods vector
meth["weight"]<-"2l.heckman"
```

Impute the missingness:
```{r eval=FALSE, cache=TRUE}
imp <- mice(
  dat, # dataset with missing values
  m = 5, # number of imputations
  maxit = 10,
  meth = meth, #imputation method vector
  pred = pred, #imputation predictors matrix
  meta_method = "reml",
  printFlag = FALSE
)
```
